Resolving a paradox of measurement uncertainty in ecological management

### 2018-08-17 manuscripts/manuscript.Rmd
here we go

---
title: "The importance of measurement uncertainty in ecological management"

@@

abstract: |
  Ecological management and decision-making typically focus on uncertainty about the future, but surprisingly little is known about how to account for uncertainty of the present: that is, the realities of having only partial or imperfect measurements.  Our two primary paradigms for handling decisions under uncertainty -- the precautionary principle and optimal control -- have so far given entirely contradictory results.  This paradox is best illustrated in the example of fisheries management, where many ideas that guide thinking about ecological decision making were first developed.  We find that simplistic optimal control approaches have repeatedly concluded that a manager should increase catch quotas when faced with greater uncertainty about the fish biomass. Current best practices take a more precautionary approach, decreasing catch quotas by a fixed amount to account for uncertainty.  Using comparisons to both simulated and historical catch data, we find that neither approach is sufficient to avoid stock collapses under moderate observational uncertainty.  Using partially observed Markov decision process (POMDP) methods, we demonstrate how this paradox arises from flaws in the standard theory, which contributes to over-exploitation of fisheries and increased probability of economic and ecological collapse.  In contrast, we find POMDP-based management avoids such over-exploitation while also generating higher economic value.  These results have significant implications for how we handle uncertainty in both fisheries and ecological management more generally.

keywords:
  - Measurement uncertainty
  - POMDP
  - Decision theory
  - Fisheries
  - Conservation

@@

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE,  dev="cairo_pdf", cache = 1)
library(tidyverse)
library(gridExtra)
```

@@

The challenge of making decisions under uncertainty is obviously not unique to fisheries, and the existing methods we critique here can be found in a wide range of population and ecosystem management challenges, including fire ecology [@Richards1999], invasive species [@Blackwood2010; @Shea1998], disease outbreaks [@Shea2014; @Li2017], and protecting biodiversity [@Dee2017; @Iacona2017].  Yet fisheries conservation and management has long been both a crucible and proving ground for the theory of ecological management more generally, including topics such as adaptive management [@Walters1978], ecosystem-based management [@Levin2008] and resilience thinking [@Holling1973; @May1977], while also giving rise to the sub-discipline of resource economics [@Beverton1957; @Schaefer1954; @Gordon1954], thanks to its global relevance, long history, and readily available data.  Despite the generality of the models and issue of measurement uncertainty considered here, any application must be grounded in both the history and context of the particular decision problem. In light of this, we begin with a brief background on existing decision theory approaches common in fisheries management.

In this paper, we examine three strategies to managing a natural population, such as a fishery, under such uncertainty.  This problem has a long history in both theory and practice, where seemingly innoccous simplifying assumptions have led to a deep paradox in how managers should respond to uncertainty.  The precautionary principle [@Kriebel2001] would seem to suggest that the greater the uncertainty in our models, the more cautious we should be in management policies.  Yet a well-known mathematical proof by @Reed1979 established that under quite general conditions, the optimal strategy under uncertainty is no different than under no uncertainty at all.  Attempts to resolve this have since only deepened this paradox, arguing that adding additional sources of uncertainty, such as imperfect measurements, should lead to optimal harvest that increases with greater uncertainty [e.g. @Clark1986; @Sethi2005].  Meanwhile, much of actual practice has (perhaps fortunately, as we shall see) ignored these results in favor of a more heuristic application of the precautionary principle.  Here, we show that this paradox can be resolved by uncovering previous shortcuts and tackling the optimal management problem head-on, which we can solve thanks to the computational efficiency of sophisticated algorithms developed more recently in the field of robotics.  This approach allows us to reject this paradox largely in favor of the precautionary principle. Further, our comparisons show that widely recognized existing approaches in both theory and practice are not by themselves sufficient to manage a population under substantial uncertainty over the long term, but our recently borrowed approach from robotics literature can under identical conditions sustain both higher biomass and higher economic yields than those approaches.  This conclusion has important implications not only for fisheries but eocsystem more broadly, underscoring the danger of not correctly accounting for measurement uncertainty when making decisions, and further to the significant opportunity to learn from fields such as robotics when faced with complex and uncertain decisions.

@@

We consider the management problem of setting catch quotas for a marine fishery in the face of imperfect information about the current stock size and uncertainty about future recruitment.  We seek to determine the sequence of actions $h_t$ for $t \in [1, ... \infty]$ that maximize the net present value (discounted sum of all future profits) of the fishery.  We will denote the discount rate $\gamma$. For simplicity we will once again follow classic theory and assume a fixed price for fish, (equivalently, measuring our value in units of discounted fish rather than discounted dollars), though again this assumption is easy to relax. Each year, the manager also obtains an estimate $y_t$ of the true $x_t$ stock size subject to some measurement uncertainty $\xi_t$

@@

For simplicity we will assume measurement uncertainty is normally distributed with standard deviation $\sigma_m$,  $\xi \sim \mathcal{N}(1, \sigma_m)$.  Given an estimated population model along with this (uncertain) measurement of the stock size, the manager must choose the set of a harvest quotas $\lbrace h_t \rbrace$ for $t \in [0, 1, \infty]$ that maximizes expected long term utility of the stock:

@@

For convenience of presentation but without loss of generality, we will assume this utility is simply equal to the harvest itself; in economic terms, a fixed price for fish without cost on effort:

@@

where the $\min$ function ensures no additional reward for attempting to harvest more than the entire stock.  Existing optimal control techniques (@Reed1979) prove these results are not dependent on the particular details of this utility.  Using this form allows a more direct comparison to approaches such as maximum sustainable yield (MSY), that do not include an economic model but merely maximize long-term catch. In principle, utility could also reflect ecosystem services or other intrinsic values for fish left in the sea and not merely the market for consumption [@Halpern2013].

@@

Maximum Sustainable Yield (MSY) remains the basis of international law (including the UN,  IWC, IATTC, ICCAT, ICNAF; @Mace2001) and a familiar standard of management in terrestrial ecosystems as well as aquatic [@Clark1990].  Critics have for some time observed the limitations of harvesting at MSY in face of uncertainty in stock sizes and population dynamics [@Larkin1977; @Botsford1997].  Many US fisheries reflect this uncertainty through a series of adjustments that effectively reduce the target fishing mortality level to reflect this uncertainty.  Typically, a stock assessment model provides a best estimate of $B_{MSY}$ and corresponding mortality $F_{MSY}$ is used to define the stock Over-Fishing Limit, (OFL). Based on this, a somewhat lower level is set as the Allowable Biological Catch (ABC), reflecting uncertainty in the stock assessment. To reflect possible uncertainty between reported and actual catch, the ABC is reduced somewhat further to define the Total Allowable Catch, (TAC), which forms the basic unit of management for many such fisheries.  To reflect this process, our analysis will also consider policies in which the harvest quota is set at 80% of the level expected under the MSY policy: $H_{TAC}(t) = 0.8 \cdot F_{MSY} \cdot B_t$, for a biomass estimated at $B(t)$ in year $t$. To distinguish this approach from MSY, we will refer to this approach as a TAC policy.  This more closely represents a heuristic [e.g. @Hilborn2010] or resilience-based approach [sensu @Fischer2009] than an optimization-based policy.  Importantly, this approach shares the fundamentally stationary assumptions of an MSY policy by defining a constant mortality rather than a dynamic policy.  Thus, despite being more cautious overall and generating lower economic yield than a constant escapement policy, TAC policies continue to harvest at non-zero rates even if a stock falls below $B_{MSY}$, while the constant escapement policy does not.

It is important to note that actual management strategies are quite diverse.  Our focus on comparisons TAC, MSY, and Constant Escapement (CE, also refereed to as MEY, see below) reflects only the subset of current practices.  Another increasingly common approach to address uncertainty in both process and models is known as Management Strategy Evaluation [MSE, @Smith1994;  @Dorner2009; @Bunnefeld2011; @Punt2016].  MSE performs forward simulations under a suite of pre-selected candidate strategies to determine which strategy achieves the best expected outcomes. Crucially, by limiting the analysis to a suite of candidate strategies over a finite number of replicate simulations, this approach is able to accommodate substantially greater complexity and uncertainty than has been possible with more formal optimal control approaches we discuss below.  Optimal control solutions rely on dynamic programming to consider all possible strategies (all possible sequences of harvest under all possible realizations of stock dynamics).  As such, MSE can be viewed as an approximation to the optimal control solutions considered below, a notion that has been formalized in the literature on approximate dynamic programming [@Nicol2011].

@@

We consider the average fish biomass across 100 replicate simulations of the same stock dynamics (Eq 4) under three different management strategies: constant escapement (CE), total allowable catch (TAC; equal to 80% MSY), and partially observed Markov decision process (POMDP) management.  Each suite of simulations is considered under three sequentially higher measurement error regimes, from $\sigma_m = 0$ (no measurement errors), $\sigma_m = 0.1$ (low error), and $\sigma_m = 0.15$ (moderate error), as indicated, while stochastic recruitment (environmental noise) is set to a moderate $\sigma_g = 0.15$.  Empirical estimates of measurement error in stock sizes vary widely, with estimates often dependent upon methodological assumptions.  Early estimates have put possible error as high as 50% [@Clark1986] and more recent analyses have suggested an average of closer to 36%, though with cases that may be much higher [@Ralston2011].  Consequently, range between 0 to 15% provides a probably conservative estimate of typical unceratinty ranges, which we show is nevertheless large enough to drive substantially negative ecological and economic outcomes under management strategies that do not appropriately account for that error over the long run.

>>>>

---
title: "Resolving a paradox of measurement uncertainty in ecological management"

@@

abstract: |
  Ecological management and decision-making typically focus on uncertainty about the future, but surprisingly little is known about how to account for uncertainty of the present: that is, the realities of having only partial or imperfect measurements.  Our primary paradigms for handling decisions under uncertainty -- the precautionary principle and optimal control -- have so far given contradictory results.  This paradox is best illustrated in the example of fisheries management, where many ideas that guide thinking about ecological decision making were first developed.  We find that simplistic optimal control approaches have repeatedly concluded that a manager should increase catch quotas when faced with greater uncertainty about the fish biomass. Current best practices take a more precautionary approach, decreasing catch quotas by a fixed amount to account for uncertainty.  Using comparisons to both simulated and historical catch data, we find that neither approach is sufficient to avoid stock collapses under moderate observational uncertainty.  Using partially observed Markov decision process (POMDP) methods, we demonstrate how this paradox arises from flaws in the standard theory, which contributes to over-exploitation of fisheries and increased probability of economic and ecological collapse.  In contrast, we find POMDP-based management avoids such over-exploitation while also generating higher economic value.  These results have significant implications for how we handle uncertainty in both fisheries and ecological management more generally.

keywords:
  - POMDP,
  - measurement uncertainty,
  - decision theory,
  - fisheries,
  - conservation

@@

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE,  dev="cairo_pdf", cache = TRUE)
library(tidyverse)
library(gridExtra)
```

@@

The challenge of making decisions under uncertainty is obviously not unique to fisheries. Existing methods we optimal control methods which similarly assume perfect measurements can be found in a wide range of population and ecosystem management challenges, including fire ecology [@Richards1999], invasive species [@Blackwood2010; @Shea1998], disease outbreaks [@Shea2014; @Li2017], and protecting biodiversity [@Dee2017; @Iacona2017], and in common textbooks and reviews [@Mangel1985; @Clark1990; @Marescot2013].  Yet fisheries conservation and management has long been both a crucible and proving ground for the theory of ecological management more generally, including topics such as adaptive management [@Walters1978], ecosystem-based management [@Levin2008] and resilience thinking [@Holling1973; @May1977], while also giving rise to the sub-discipline of resource economics [@Beverton1957; @Schaefer1954; @Gordon1954], thanks to its global relevance, long history, and readily available data.  Despite the generality of the models and issue of measurement uncertainty considered here, any application must be grounded in both the history and context of the particular decision problem. In light of this, we begin with a brief background on existing decision theory approaches common in fisheries management.

In this paper, we examine three strategies to managing a natural population, such as a fishery, under such uncertainty.  This problem has a long history in both theory and practice, where seemingly innocuous simplifying assumptions have led to a deep paradox in how managers should respond to uncertainty.  The precautionary principle [@Kriebel2001] would seem to suggest that the greater the uncertainty in our models, the more cautious we should be in management policies.  Yet a well-known mathematical proof by @Reed1979 established that under quite general conditions, the optimal strategy under uncertainty is no different than under no uncertainty at all.  Attempts to resolve this have since only deepened this paradox, arguing that adding additional sources of uncertainty, such as imperfect measurements, should lead to optimal harvest that increases with greater uncertainty [e.g. @Clark1986; @Sethi2005].  Meanwhile, much of actual practice has (perhaps fortunately, as we shall see) ignored these results in favor of a more heuristic application of the precautionary principle.  Here, we show that this paradox can be resolved by uncovering previous shortcuts and tackling the optimal management problem head-on, which we can solve thanks to the computational efficiency of sophisticated algorithms developed more recently in the field of robotics.  This approach allows us to reject this paradox largely in favor of the precautionary principle. Further, our comparisons show that widely recognized existing approaches in both theory and practice are not by themselves sufficient to manage a population under substantial uncertainty over the long term, but our recently borrowed approach from robotics literature can under identical conditions sustain both higher biomass and higher economic yields than those approaches.  This conclusion has important implications not only for fisheries but ecosystem more broadly: underscoring the danger of not correctly accounting for measurement uncertainty when making decisions, and further to the significant opportunity to learn from fields such as robotics when faced with complex and uncertain decisions.

@@

We consider the management problem of setting catch quotas for a marine fishery in the face of imperfect information about the current stock size and uncertainty about future recruitment.  We seek to determine the sequence of actions $h_t$ for $t \in [1, ... \infty]$ that maximize the net present value (discounted sum of all future profits) of the fishery.  We will denote the discount rate $\gamma$. For simplicity we will once again follow classic theory and assume a fixed price for fish, (equivalently, measuring our value in units of discounted fish rather than discounted dollars). Each year, the manager also obtains an estimate $y_t$ of the true $x_t$ stock size subject to some measurement uncertainty $\xi_t$

@@

For simplicity, we will assume measurement uncertainty is normally distributed with standard deviation $\sigma_m$,  $\xi \sim \mathcal{N}(1, \sigma_m)$.  Given an estimated population model along with this (uncertain) measurement of the stock size, the manager must choose the set of a harvest quotas $\lbrace h_t \rbrace$ for $t \in [0, 1, \infty]$ that maximizes expected long term utility of the stock:

@@

For convenience of presentation, we will assume this utility is simply equal to the harvest itself.

@@

where the $\min$ function ensures no additional reward for attempting to harvest more than the entire stock.  In economic terms, the corresponds to a fixed price per fish caught, with no proportional cost on effort required to do so, but the results do not depend on this assumption, as existing analyses have previously established [@Reed1979].  Using this form allows a more direct comparison to approaches such as maximum sustainable yield (MSY) which do not include an economic model but merely maximize long-term catch. In principle, utility could also reflect ecosystem services or other intrinsic values for fish left in the sea and not merely the market for consumption [@Halpern2013].

@@

Maximum Sustainable Yield (MSY) remains the basis of international law (including the UN,  IWC, IATTC, ICCAT, ICNAF; @Mace2001) and a familiar standard of management in terrestrial ecosystems as well as aquatic [@Clark1990].  Critics have for some time observed the limitations of harvesting at MSY in face of uncertainty in stock sizes and population dynamics [@Larkin1977; @Botsford1997].  Many US fisheries reflect this uncertainty through a series of adjustments that effectively reduce the target fishing mortality level to reflect this uncertainty.  Typically, a stock assessment model provides a best estimate of $B_{MSY}$ and corresponding mortality $F_{MSY}$ is used to define the stock Over-Fishing Limit, (OFL). Based on this, a somewhat lower level is set as the Allowable Biological Catch (ABC), reflecting uncertainty in the stock assessment. To reflect possible uncertainty between reported and actual catch, the ABC may be reduced further to define the Total Allowable Catch, (TAC), which forms the basic unit of management for many such fisheries.  To reflect this process, our analysis will also consider policies in which the harvest quota is set at 80% of the level expected under the MSY policy: $H_{TAC}(t) = 0.8 \cdot F_{MSY} \cdot B_t$, for a biomass estimated at $B(t)$ in year $t$. To distinguish this approach from MSY, we will refer to this approach as a TAC policy.  This more closely represents a heuristic [e.g. @Hilborn2010] or resilience-based approach [sensu @Fischer2009] than an optimization-based policy.  Importantly, this approach shares the fundamentally stationary assumptions of an MSY policy by defining a constant mortality rather than a dynamic policy.  Thus, despite being more cautious overall and generating lower economic yield than a constant escapement policy, TAC policies continue to harvest at non-zero rates even if a stock falls below $B_{MSY}$, while the constant escapement policy does not.

It is important to note that actual management strategies are quite diverse.  Our focus on comparisons TAC, MSY, and Constant Escapement (CE) reflects only the subset of current practices.  Another increasingly common approach to address uncertainty in both process and models is known as Management Strategy Evaluation [MSE, @Smith1994;  @Dorner2009; @Bunnefeld2011; @Punt2016].  MSE performs forward simulations under a suite of pre-selected candidate strategies to determine which strategy achieves the best expected outcomes. Crucially, by limiting the analysis to a suite of candidate strategies over a finite number of replicate simulations, this approach is able to accommodate substantially greater complexity and uncertainty than has been possible with more formal optimal control approaches we discuss below.  Optimal control solutions rely on dynamic programming to consider all possible strategies (all possible sequences of harvest under all possible realizations of stock dynamics).  As such, MSE can be viewed as an approximation to the optimal control solutions considered below, a notion that has been formalized in the literature on approximate dynamic programming [@Nicol2011]. Future work may be able to leverage the forward-simulation approach used in MSE to extend the POMDP methods introduced here to yet more complex dynamics.

@@

We consider the average fish biomass across 100 replicate simulations of the same stock dynamics (Eq 4) under three different management strategies: constant escapement (CE), total allowable catch (TAC; equal to 80% MSY), and partially observed Markov decision process (POMDP) management.  Each suite of simulations is considered under three sequentially higher measurement error regimes, from $\sigma_m = 0$ (no measurement errors), $\sigma_m = 0.1$ (low error), and $\sigma_m = 0.15$ (moderate error), as indicated, while stochastic recruitment (environmental noise) is set to a moderate $\sigma_g = 0.15$.  Empirical estimates of measurement error in stock sizes vary widely, with estimates often dependent upon methodological assumptions.  Early estimates have put possible error as high as 50% [@Clark1986] and more recent analyses have suggested an average of closer to 36%, though with cases that may be much higher [@Ralston2011].  Consequently, range between 0 to 15% provides a probably conservative estimate of typical uncertainty ranges, which we show is nevertheless large enough to drive substantially negative ecological and economic outcomes under management strategies that do not appropriately account for that error over the long run.
